---
title: "Metadata compilation figure"
date: 2026-09-02
format:
  html:
    toc: true
editor: source
---

# Overview

Make multi-panel metadata figure for meta-eukomic effort.

| Panel | Contents |
|------------------------------|------------------------------------------|
| `heat_sample` | Filter, depth, lysis reagent, physical disruption |
| `bead_size` | Bead size range (mm) |
| `heat_chemistry` | Lysis timing/temperature, extraction kit, DNase, elution, quantification |
| `rna_qc` | RIN, ng/µl, total µg |
| `heat_sequencing` | Library prep, sequencer, sequencing facility |
| `assembly_stats` | Reads, BUSCO, transcript and assembly metrics |

color guide:

- **Colored** -- method was used
- **Grey** -- method was recorded as unknown.
- **White** -- no record / method not used / NA

# Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(scales)
library(patchwork)
library(ggnewscale)

# size aprameters for the output figure. 
cell_mm <- 3.2
show_legends <- FALSE #no legends if FALSE

# header strips need ggh4x for exact panel sizing used later.
has_ggh4x <- requireNamespace("ggh4x", quietly = TRUE)

if (!has_ggh4x) {
  message("ggh4x not installed - group header strips will be disabled. ",
          "install.packages('ggh4x')")
}
```

# Import and check metadata

`clean_names()` standardises column names; the two checks below report duplicated sample IDs and all-empty columns before anything downstream depends on them.


> SH To do, make the input file anonymous for labs. Create look up table.

```{r}
#| label: import-metadata
#| message: false

metadata_raw <- read_csv(
  "RNAExtractions_Metadata_V1.1.csv",
  show_col_types = FALSE
) |>
  clean_names() |> 
  mutate(sequencing_facility_raw = replace_na(sequencing_facility_raw, "UCSD"))

# Check uniquies. one row per sequence_id.
dupe_ids <- metadata_raw |> get_dupes(sequence_id)

if (nrow(dupe_ids) > 0) {
  warning("Duplicated sequence_id values found:")
  print(dupe_ids)
} else {
  cat("No duplicated sequence_id values.\n")
}

# KEEP empty columns so all parameters line up
empty_cols <- setdiff(
  names(metadata_raw),
  names(remove_empty(metadata_raw, which = "cols"))
)

cat("All-NA columns:", if (length(empty_cols)) paste(empty_cols, collapse = ", ") else "none", "\n")

cat("Rows:", nrow(metadata_raw), " Columns:", ncol(metadata_raw), "\n")
```

## Depth

`sequencing_facility` now comes straight from `RNAExtractions_Metadata_V1_cleaned.csv` (resolved once, in `scripts/clean_rna_extraction_metadata.R`, rather than by a join here -- see `facility_filled` in that file for which rows were backfilled rather than read from the raw sheet).

```{r}
#| label: derive-depth
#| message: false

stopifnot(!anyNA(metadata_raw$sequencing_facility))

metadata_clean <- metadata_raw |>
  mutate(depth = if_else(filter %in% c("A", "B"), "30 m", "35 m"))
```

## Code categorical levels

Per-column order, naming, unchanged in meaning from the original. Comma-delimited values are split into separate heatmap columns further down.

```{r}
#| label: recode-levels

metadata_recoded <- metadata_clean |>
  mutate(
    lysis_reagent = str_replace_all(lysis_reagent, "none", "Unknown_lysis"),
    physical_disruption_method = replace_na(physical_disruption_method, "Unknown_disrupt"),
    physical_disruption_method = str_replace_all(
      physical_disruption_method, "Vortex with beads", "Vortex,Beads"
    ),
    dnase_brand_clean = case_when(
      str_detect(dnase_brand_clean, "Kit-provided") ~ "Kit-provided",
      dnase_brand_clean == "Qiagen + Turbo DNase"   ~ "Qiagen RNase-free DNase,Turbo DNase",
      is.na(dnase_brand_clean)                      ~ "Unknown_dnase",
      .default = dnase_brand_clean
    ),
    across(c(bead_beat_time_min, lysis_temp_c, lysis_incubation_time_min), as.numeric)
  )

sample_levels <- sort(unique(metadata_recoded$sequence_id))

# View(metadata_recoded)
```

```{r}
# str(metadata_recoded)
lab_lookup <- as.data.frame(metadata_recoded) |>
dplyr::select(lab, lab_id)
write.csv(lab_lookup, file = "output-files/INTERNAL-lablookup.csv")
metadata_recoded <- metadata_recoded |> dplyr::select(-lab)
write.csv(metadata_recoded, "metadata_09032026.csv")
```

```{r}
metadata_recoded <- read.csv("metadata_09032026.csv")
```

# Variable key

One row per heatmap column. This table controls **order, grouping, labelling and colour** for every column in the three heatmap panels -- it replaces the paired `variable_fill` / `scale_fill` lists, which had to stay the same length and in the same order by hand.

- `type = "presence"` -- column is one level of `source`; filled with the block colour when that level is present for a sample.
- `type = "own"` -- column carries its own values and gets its own fill scale (see `own_scales` below).
- `is_unknown = TRUE` -- filled grey rather than the block colour.

```{r}
#| label: variable-key

var_key <- tribble(
  ~panel, ~block, ~label, ~source, ~level, ~type, ~is_unknown,
  #panel 1: sample + lysis + disruption
  "sample", "sample", "Filter", "filter", NA, "own", FALSE,
"sample","sample", "Depth","depth",NA, "own",FALSE,
"sample","lysis", "β-mercaptoethanol","lysis_reagent","beta-mercaptoethanol","presence",FALSE,
"sample","lysis", "+TCEP", "lysis_reagent","TCEP","presence",FALSE,
"sample","lysis", "Unknown lysis buffer","lysis_reagent","Unknown_lysis", "presence",TRUE,
"sample","disruption","Beads", "physical_disruption_method", "Beads", "presence",FALSE,
"sample","disruption","Vortex","physical_disruption_method", "Vortex","presence",FALSE,
"sample","disruption","Unknown method","physical_disruption_method", "Unknown_disrupt", "presence",TRUE,
#panel 2: lysis conditions + extraction chemistry ----------------------
"chemistry", "timing","Bead beat (min)", "bead_beat_time_min", NA,"own", FALSE,
"chemistry", "timing","Lysis incubation (min)","lysis_incubation_time_min",NA,"own", FALSE,
"chemistry", "timing","Lysis temp °C","lysis_temp_c", NA,"own", FALSE,
"chemistry", "extraction","NucleoMag RNA", "extraction_protocol","NucleoMag RNA", "presence",FALSE,
"chemistry", "extraction","NucleoSpin RNA","extraction_protocol","NucleoSpin RNA","presence",FALSE,
"chemistry", "extraction","Qiagen AllPrep RNA/DNA","extraction_protocol","Qiagen AllPrep RNA/DNA","presence",FALSE,
"chemistry", "extraction","Qiagen PowerWater RNA", "extraction_protocol","Qiagen PowerWater RNA", "presence",FALSE,
"chemistry", "extraction","Qiagen RNeasy", "extraction_protocol","Qiagen RNeasy", "presence",FALSE,
"chemistry", "extraction","Qiagen RNeasy Plant Mini Kit","extraction_protocol","Qiagen RNeasy Plant Mini Kit","presence",FALSE,
"chemistry", "extraction","RNAqueous-4PCR Total RNA Isolation","extraction_protocol","RNAqueous-4PCR Total RNA Isolation","presence",FALSE,
"chemistry", "extraction","Thermo Fisher Trizol","extraction_protocol","Thermo Fisher Trizol","presence",FALSE,
"chemistry", "extraction","Zymo Quick DNA/RNA miniprep kit", "extraction_protocol","Zymo Quick DNA/RNA miniprep kit", "presence",FALSE,
"chemistry", "extraction","Zymo RNA MiniPrep Plus","extraction_protocol","Zymo RNA MiniPrep Plus","presence",FALSE,
"chemistry", "extraction","Zymo RNA Quick","extraction_protocol","Zymo RNA Quick","presence",FALSE,
"chemistry", "dnase", "Kit-provided","dnase_brand_clean","Kit-provided","presence",FALSE,
"chemistry", "dnase", "Qiagen RNase-free DNase", "dnase_brand_clean","Qiagen RNase-free DNase", "presence",FALSE,
"chemistry", "dnase", "Turbo DNase", "dnase_brand_clean","Turbo DNase", "presence",FALSE,
"chemistry", "dnase", "Unknown DNase", "dnase_brand_clean","Unknown_dnase", "presence",TRUE,
"chemistry", "elution", "RNase-free water","elution_solution", "RNase-free water","presence",FALSE,
"chemistry", "elution", "Kit elution solution","elution_solution", "Elution solution","presence",FALSE,
"chemistry", "quant", "Bioanalyzer", "quantification_method","Bioanalyzer", "presence",FALSE,
"chemistry", "quant", "Qubit", "quantification_method","Qubit", "presence",FALSE,
"chemistry", "quant", "RiboGreen", "quantification_method","RiboGreen", "presence",FALSE,
"chemistry", "quant", "Nanodrop","quantification_method","Nanodrop","presence",FALSE,
#panel 3: library prep + sequencing ------------------------------------
"sequencing","library", "PolyA", "library_prep_type","PolyA", "presence",FALSE,
"sequencing","library", "rRNA depletion","library_prep_type","rRNA depletion","presence",FALSE,
"sequencing","sequencer", "Illumina NovaSeq 6000", "sequencing_machine", "Illumina NovaSeq 6000", "presence",FALSE,
"sequencing","sequencer", "Illumina NextSeq 2000 Flow Cell P2","sequencing_machine", "Illumina NextSeq 2000 Flow Cell P2","presence",FALSE,
"sequencing","sequencer", "Illumina NovaSeq S4", "sequencing_machine", "Illumina NovaSeq S4", "presence",FALSE,
"sequencing","sequencer", "Illumina HiSeq 4000", "sequencing_machine", "Illumina HiSeq 4000", "presence",FALSE,
"sequencing","sequencer", "Illumina NovaSeq S2", "sequencing_machine", "Illumina NovaSeq S2", "presence",FALSE,
"sequencing","sequencer", "NovaSeq X (PE150)", "sequencing_machine", "NovaSeq X (PE150)", "presence",FALSE,
"sequencing","facility","Columbia Genome Center","sequencing_facility","Columbia Genome Center","presence",FALSE,
"sequencing","facility","UGA", "sequencing_facility","UGA", "presence",FALSE,
"sequencing","facility","UCSD","sequencing_facility","UCSD","presence",FALSE,
"sequencing","facility","Genewiz", "sequencing_facility","Genewiz", "presence",FALSE,
"sequencing","facility","Northwest Genomics Center", "sequencing_facility","Northwest Genomics Center", "presence",FALSE
) |>
  mutate(
    panel     = factor(panel, levels = c("sample", "chemistry", "sequencing")),
    block     = factor(block, levels = unique(block)),
    label     = factor(label, levels = unique(label)),
    fill_key  = if_else(is_unknown, "unknown", as.character(block))
  )

stopifnot(
  !anyDuplicated(var_key$label),
  all(var_key$source %in% names(metadata_recoded))
)

var_key |> count(panel, block) |> print(n = Inf)
```

## Colours

One colour per variable block, plus grey for unknowns. These are the same hex values as the original figure.

```{r}
#| label: palettes

block_pal <- c(
  lysis      = "#b55050",
  disruption = "#b57550",
  extraction = "#e3a02d",
  dnase      = "#4b5c13",
  elution    = "#1a6e20",
  quant      = "#1a6e4a",
  library    = "#3468a8",
  sequencer  = "#6134a8",
  facility   = "#863594",
  unknown    = "#9e9e9e"
)

# Group header text for the facet strips.
block_labels <- c(
  sample     = "Sample",
  lysis      = "Lysis reagent",
  disruption = "Disruption",
  timing     = "Lysis conditions",
  extraction = "Extraction kit",
  dnase      = "DNase",
  elution    = "Elution",
  quant      = "Quantification",
  library    = "Library prep",
  sequencer  = "Sequencer",
  facility   = "Sequencing facility"
)

# Self-scaled columns. Stored as functions so each plot builds a fresh,
# untrained scale object.
guide_or_none <- function() if (show_legends) "colourbar" else "none"

own_scales <- list(
  filter = function() {
    scale_fill_manual(
      values = c(A = "#ae2f6d", B = "#67c9ba", C = "#fbef52",
                 D = "#f2bfc9", E = "#cc504a", F = "#3f65c5"),
      na.value = "white", name = "Filter",
      guide = if (show_legends) "legend" else "none"
    )
  },
  depth = function() {
    scale_fill_manual(
      values = c("30 m" = "#aeb7e0", "35 m" = "#333e80"),
      na.value = "white", name = "Depth",
      guide = if (show_legends) "legend" else "none"
    )
  },
  bead_beat_time_min = function() {
    scale_fill_stepsn(
      colors = c("#d6d0d4", "#d4b6cb", "#cf99be", "#c75da7", "#c22593"),
      na.value = "white", name = "Bead beat (min)", guide = guide_or_none()
    )
  },
  lysis_incubation_time_min = function() {
    scale_fill_stepsn(
      colors = c("#d6d0d4", "#d4b6cb", "#cf99be", "#c75da7", "#c22593"),
      na.value = "white", name = "Lysis incubation (min)", guide = guide_or_none()
    )
  },
  lysis_temp_c = function() {
    scale_fill_stepsn(
      colors = c("#ebdd5e", "#d4c220", "#d4ad20", "#bd821c",
                 "#bd671c", "#bd4a1c", "#bd241c"),
      breaks = c(4, 20, 30, 40, 50, 65),
      na.value = "white", name = "Lysis temp °C", guide = guide_or_none()
    )
  }
)

stopifnot(all(var_key$source[var_key$type == "own"] %in% names(own_scales)))
```

# Tile data

Long format throughout. A level that is present becomes a row; anything absent simply has no row and stays white.

```{r}
#| label: tile-data

presence_key <- var_key |> filter(type == "presence")

presence_long <- metadata_recoded |>
  select(sequence_id, all_of(unique(presence_key$source))) |>
  pivot_longer(-sequence_id, names_to = "source", values_to = "level") |>
  separate_longer_delim(level, delim = ",") |>
  mutate(level = str_trim(level)) |>
  filter(!is.na(level), level != "")

# Anything in the data that the key does not describe would silently vanish.
# Report it instead.
unmatched <- presence_long |>
  anti_join(presence_key, by = c("source", "level")) |>
  distinct(source, level)
if (nrow(unmatched) > 0) {
  warning("Levels present in the data but missing from var_key:")
  print(unmatched)
} else {
  cat("All observed levels are described in var_key.\n")
}

presence_tiles <- presence_long |>
  inner_join(presence_key, by = c("source", "level")) |>
  mutate(sequence_id = factor(sequence_id, levels = sample_levels))
```

# Plot helpers

## Themes

```{r}
#| label: theme-helpers

theme_panel_base <- function(y_labels = FALSE) {
  theme_void() +
    theme(
      axis.text.y      = if (y_labels) {
                           element_text(colour = "black", size = 6, hjust = 1)
                         } else {
                           element_blank()
                         },
      axis.title       = element_blank(),
      legend.title     = element_blank(),
      legend.position  = if (show_legends) "right" else "none",
      panel.border     = element_rect(fill = NA, linewidth = 0.8, colour = "black"),
      panel.background = element_rect(colour = NA, fill = "transparent"),
      plot.background  = element_rect(colour = NA, fill = "transparent"),
      plot.margin      = margin(0, 0, 0, 0, "cm"),
      panel.spacing    = unit(0.06, "cm")
    )
}

theme_heat <- function(y_labels = FALSE) {
  theme_panel_base(y_labels = y_labels) +
    theme(
      axis.text.x      = element_text(size = 7, angle = 90, hjust = 0),
      strip.background = element_blank(),
      strip.text.x     = element_text(size = 7, face = "bold", margin = margin(b = 2))
    )
}

theme_dots <- function(strip = TRUE, y_labels = FALSE) {
  theme_panel_base(y_labels = y_labels) +
    theme(
      axis.text.x      = element_text(size = 5, angle = 45, hjust = 1, vjust = 1),
      axis.ticks.x     = element_line(linewidth = 0.4, colour = "grey80"),
      panel.grid.major = element_line(linetype = 3, linewidth = 0.4, colour = "grey80"),
      panel.grid.minor = element_line(linetype = 3, linewidth = 0.4, colour = "grey80"),
      strip.background = element_blank(),
      strip.text.x     = if (strip) element_text(size = 7) else element_blank()
    )
}
```

## Square cells and group headers

`ggh4x::force_panelsizes()` pins each panel to an exact size, so cells stay square no matter how patchwork allocates width. It is also what makes the grouped layout possible: `coord_fixed()` cannot be combined with `space = "free_x"`, so without ggh4x the heatmaps drop the group header strips and fall back to `coord_fixed()` -- square cells, no headers, original look.

```{r}
#| label: square-cells

layout_heatmap <- function(n_cols_by_facet, cell = cell_mm) {
  if (has_ggh4x) {
    list(
      facet_grid(
        cols     = vars(block),
        scales   = "free_x",
        space    = "free_x",
        labeller = labeller(block = block_labels)
      ),
      ggh4x::force_panelsizes(
        rows = unit(length(sample_levels) * cell, "mm"),
        cols = unit(n_cols_by_facet * cell, "mm")
      )
    )
  } else {
    coord_fixed(ratio = 1)
  }
}
```

## Heatmap builder

Builds one heatmap panel. Layer order: an empty outlined grid (so every cell has a border, including white ones), then the block-coloured presence tiles, then one layer per self-scaled column.

```{r}
#| label: heatmap-builder

build_heatmap <- function(panel_id, y_labels = FALSE) {
  key <- var_key |> filter(panel == panel_id) |> droplevels()

  grid_cells <- expand_grid(
    sequence_id = factor(sample_levels, levels = sample_levels),
    label       = key$label
  ) |>
    left_join(key |> select(label, block), by = "label")

  tiles <- presence_tiles |>
    filter(label %in% key$label) |>
    mutate(
      label = factor(as.character(label), levels = levels(key$label)),
      block = factor(as.character(block), levels = levels(key$block))
    )

  p <- ggplot(mapping = aes(x = label, y = sequence_id)) +
    geom_tile(data = grid_cells, fill = NA,
              colour = "grey20", linewidth = 0.2, width = 0.95) +
    geom_tile(data = tiles, aes(fill = fill_key),
              colour = "grey20", linewidth = 0.2, width = 0.95) +
    scale_fill_manual(values = block_pal, guide = "none")

  own_rows <- which(key$type == "own")
  for (i in own_rows) {
    p <- p + new_scale_fill()
    own_data <- tibble(
      sequence_id = factor(metadata_recoded$sequence_id, levels = sample_levels),
      label       = factor(as.character(key$label[i]), levels = levels(key$label)),
      block       = factor(as.character(key$block[i]), levels = levels(key$block)),
      value       = metadata_recoded[[key$source[i]]]
    )
    p <- p +
      geom_tile(data = own_data, aes(fill = value),
                colour = "grey20", linewidth = 0.2, width = 0.95) +
      own_scales[[key$source[i]]]()
  }

  n_cols <- key |> count(block) |> pull(n)

  p +
    layout_heatmap(n_cols) +
    scale_x_discrete(position = "top", expand = expansion(0, 0)) +
    scale_y_discrete(expand = expansion(0, 0)) +
    theme_heat(y_labels = y_labels)
}
```

# Heatmap panels

```{r}
#| label: fig-heat-sample
#| fig-width: 4
#| fig-height: 7

heat_sample <- build_heatmap("sample", y_labels = TRUE)
heat_sample
```

```{r}
#| label: fig-heat-chemistry
#| fig-width: 9
#| fig-height: 7

heat_chemistry <- build_heatmap("chemistry")
heat_chemistry
```

```{r}
#| label: fig-heat-sequencing
#| fig-width: 6
#| fig-height: 7

heat_sequencing <- build_heatmap("sequencing")
heat_sequencing
```

# Bead size range

```{r}
#| label: fig-bead-size
#| fig-width: 2
#| fig-height: 7
#| warning: false

bead_size <- metadata_recoded |>
  select(sequence_id, bead_size_raw, bead_size_min_mm, bead_size_max_mm) |>
  separate_longer_delim(bead_size_raw, delim = ", ") |>
  separate_longer_delim(bead_size_raw, delim = " and ") |>
  mutate(
    bead_size_raw = str_remove_all(bead_size_raw, "\\s*mm"),
    bead_size_raw = na_if(str_trim(bead_size_raw), ""),
    has_range     = str_detect(replace_na(bead_size_raw, ""), "-"),
    bead_size_min_mm = if_else(has_range, as.numeric(bead_size_min_mm), NA_real_),
    bead_size_max_mm = if_else(has_range, as.numeric(bead_size_max_mm), NA_real_)
  ) |>
  separate_longer_delim(bead_size_raw, delim = "-") |>
  mutate(
    bead_size_raw = as.numeric(bead_size_raw),
    sequence_id   = factor(sequence_id, levels = sample_levels)
  ) |>
  ggplot(aes(x = bead_size_raw, y = sequence_id)) +
  geom_linerange(aes(xmin = bead_size_min_mm, xmax = bead_size_max_mm),
                 linewidth = 0.5, colour = "grey30") +
  geom_point(shape = 21, size = 2.5, colour = "black", fill = "#3f4d4a") +
  scale_x_continuous(expand = expansion(mult = c(0.1, 0.1))) +
  scale_y_discrete(expand = expansion(0, 0)) +
  theme_dots(strip = FALSE)

bead_size
```

# RNA yield and quality

```{r}
#| label: fig-rna-qc
#| fig-width: 4
#| fig-height: 7
#| warning: false

rna_qc <- metadata_recoded |>
  select(sequence_id, rin, rna_conc_ng_ul, total_rna_yield_ug) |>
  pivot_longer(-sequence_id, names_to = "metric", values_to = "value_raw") |>
  mutate(
    value        = as.numeric(value_raw),
    sequence_id  = factor(sequence_id, levels = sample_levels),
    metric_label = factor(
      metric,
      levels = c("rin", "rna_conc_ng_ul", "total_rna_yield_ug"),
      labels = c("RIN", "ng/µl", "Total µg")
    )
  ) |>
  ggplot(aes(y = sequence_id)) +
  geom_linerange(aes(xmin = 0, xmax = value), linewidth = 0.5, colour = "grey30") +
  geom_point(aes(x = value, fill = metric_label),
             shape = 21, size = 2.5, colour = "black") +
  facet_grid(cols = vars(metric_label), scales = "free_x") +
  scale_fill_manual(values = c("#d94b2b", "#d9712b", "#d9912b"), guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.25))) +
  scale_y_discrete(expand = expansion(0, 0)) +
  theme_dots()

rna_qc
```

# Assembly and transcript output

```{r}
#| label: fig-assembly-stats
#| fig-width: 12
#| fig-height: 7
#| warning: false

assembly_metrics <- c(
  sequence_reads        = "Total reads",
  busco_partial         = "BUSCO partial",
  busco_complete        = "BUSCO complete",
  transcripts           = "Total transcripts",
  avg_transcript_length = "Mean transcript length (bp)",
  n_genes_genemark      = "Number of genes (GeneMark-ST)",
  longest_transcript    = "Longest transcript",
  transcript_n50        = "N50",
  transcripts_500bp     = "Transcripts > 500 bp",
  transcripts_1000bp    = "Transcripts > 1000 bp",
  total_assembly_length = "Total assembly length (bp)"
)

assembly_pal <- c("#9AC1DC", "#E1BD9E", "#D2BF8E", "#167CBA", "#B4C4D4",
                  "#6F89A5", "#A9DEEC", "#88A10B", "#A4AE35", "#7B8B75", "#6EC076")

assembly_stats <- metadata_recoded |>
  select(sequence_id, all_of(names(assembly_metrics))) |>
  pivot_longer(-sequence_id, names_to = "metric", values_to = "value_raw") |>
  mutate(
    value        = as.numeric(value_raw),
    sequence_id  = factor(sequence_id, levels = sample_levels),
    metric_label = factor(metric,
                          levels = names(assembly_metrics),
                          labels = str_wrap(unname(assembly_metrics), width = 14))
  ) |>
  ggplot(aes(y = sequence_id)) +
  geom_linerange(aes(xmin = 0, xmax = value), linewidth = 0.5, colour = "grey30") +
  geom_point(aes(x = value, fill = metric_label),
             shape = 21, size = 2.5, colour = "black") +
  facet_grid(cols = vars(metric_label), scales = "free_x") +
  scale_fill_manual(values = assembly_pal, guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.25))) +
  scale_y_discrete(expand = expansion(0, 0)) +
  theme_dots()

assembly_stats
```

# Combine

Panel widths are derived from the number of heatmap columns in each block, so they stay correct if `var_key` changes.

```{r}
#| label: fig-metadata-compilation
#| fig-width: 22
#| fig-height: 8
#| warning: false

heat_widths <- var_key |> count(panel) |> pull(n)   # sample, chemistry, sequencing

metadata_figure <-
  heat_sample +
  bead_size +
  heat_chemistry +
  rna_qc +
  heat_sequencing +
  assembly_stats +
  plot_layout(
    nrow   = 1,
    widths = c(heat_widths[1], 2.5, heat_widths[2], 9, heat_widths[3], 36),
    guides = "collect"
  )

metadata_figure
```

```{r}
#| label: save-figure
#| eval: false

# ggsave("heatmap-metadata_2026-09-01.svg", metadata_figure,
#        width = 22, height = 8, device = "svg", limitsize = FALSE)
```

# Notes

- Clustering of samples by processing method is in `cluster_samples_by_processing.qmd`.
- To add a variable to the heatmap: add one row to `var_key`. If it needs a colour of its own, add an entry to `block_pal` (and `block_labels`); if it needs its own scale, add a function to `own_scales`.

# Session information

```{r}
#| label: session-info

sessionInfo()
```
